# Reproducible Research: Peer Assessment 1
The following libraries are essential to execute the scripts in this file:
```{r, echo=TRUE}
library(ggplot2)
```


## Loading and preprocessing the data
First, we unzip the data, and read the data in a dataframe.
```{r loadData, echo=TRUE}
unzip('activity.zip')
data <- read.csv('activity.csv')
```

Second, we need to convert the date column from factor to date.
```{r convertDate, echo=TRUE}
data$date <- as.Date(data$date,"%Y-%m-%d")
```

Finally, we inspect the data, and make sure everything looks OK.
```{r insepctData, echo=TRUE}
names(data)
head(data)
tail(data)
str(data)
summary(data)
```

## What is mean total number of steps taken per day?
Make a histogram of the total number of steps taken each day.
```{r histoStepsPerDay, echo=TRUE}
steps_per_day <- aggregate(steps ~ date, data, sum)
colnames(steps_per_day) <- c("date", "steps")
p <- qplot(data=steps_per_day, x=steps, geom='bar', alpha=I(.8), binwidth=1000,
      main='Number of Steps taken per day')
p + theme(axis.title=element_text(face="bold.italic", 
   size="12", color="brown"), legend.position="top")
```
Calculate and report the mean and median total number of steps taken
per day.
```{r meanStepsPerDay, echo=TRUE, results='hide'}
meanSteps <- round(mean(steps_per_day$steps))
medianSteps <- median(steps_per_day$steps)
```

For the total number of steps taken per day, the mean is `r paste(meanSteps)` and the median is `r paste(medianSteps)`.

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis).

```{r timeseriesPlot, echo=TRUE}
steps_per_interval <- aggregate(steps ~ interval, data, sum, na.rm=T)
colnames(steps_per_interval) <- c("interval", "steps")
p <- qplot(data=steps_per_interval, x=interval, y=steps, geom='line', alpha=I(.8), 
      main='Number of Steps taken per interval')
p + theme(axis.title=element_text(face="bold.italic", 
   size="12", color="brown"), legend.position="top")

```

Which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps?
```{r maxStepsperInterval, echo=TRUE}
max_steps <- max(steps_per_interval$steps)
max_interval <- which.max(steps_per_interval$steps)
```

The interval that has the maximum steps is interval: `r steps_per_interval$interval[max_interval]` with `r max_steps` steps.

## Imputing missing values
Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)
```{r numberOfMissingValues, echo=TRUE}
sum(is.na(data$steps))
```

Make a new dataframe from data, and fill in zeroes for missing values:
```{r fillMissingSteps, echo=TRUE}
imputed <- data.frame(data)
imputed$steps[is.na(data$steps)] <- 0
```

Make a histogram of the total number of steps taken each day from the imputed data frame.
```{r histoStepsPerDayImputed, echo=TRUE}
steps_per_day_imputed <- aggregate(steps ~ date, imputed, sum)
colnames(steps_per_day_imputed) <- c("date", "steps")
p <- qplot(data=steps_per_day_imputed, x=steps, geom='bar', alpha=I(.8), binwidth=1000,
      main='Number of Steps taken per day')
p + theme(axis.title=element_text(face="bold.italic", 
   size="12", color="brown"), legend.position="top")
```
Calculate and report the mean and median total number of steps taken
per day.
```{r meanStepsPerDayImputed, echo=TRUE}
meanStepsImputed <- round(mean(steps_per_day_imputed$steps))
medianStepsImputed <- median(steps_per_day_imputed$steps)
```

For the total number of steps taken per day from the imputed dataframe, the mean is `r paste(meanStepsImputed)` and the median is `r paste(medianStepsImputed)`. The imputation strategy used shifts the distribution towards lower bounds.


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday”
and “weekend” indicating whether a given date is a weekday or weekend
day. Make a panel plot containing a time series plot (i.e. type = "l") of the
5-minute interval (x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis). The plot should look
something like the following, which was creating using simulated data:

```{r plotWeekend, echo=TRUE}
imputed$day_type <- as.factor(ifelse(weekdays(imputed$date) %in% c("Saturday", "Sunday"),
                           'weekend', 'weekday'))
 plot_data <- aggregate(steps ~ day_type + interval, imputed, mean)

p <- qplot(data=plot_data, x=interval, y=steps, facets=day_type~., geom='line', alpha=I(.8), 
      main='Number of Steps taken per interval')
p + theme(axis.title=element_text(face="bold.italic", 
   size="12", color="brown"), legend.position="top")

```
The data spikes in the mornings of the weekdays, obviously indicating that the subject is going to work. The weekend data is more adhoc.